Stability in human interaction networks: primitive typology of vertex, prominance of measures and activity statistics

### 2014-12-24 paper.tex
adicionado pdf compilado
feito retoque geral do artigo

\author{Luciano da Fontoura Costa}
  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
  \email{ldfcosta@gmail.com}
  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\

\author{Osvaldo N. Oliveira Jr.}
  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
  \email{chu@ifsc.usp.br}
 \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\

@@

\begin{abstract}
    This article reports a minimal and general characterization of interaction networks evolution. Such a task involves a selection of aspects to investigate, which lead to: 1) activity distribution in time and among participants, 2) a sound and stable classification of vertex: peripheral, intermediary and hub sectors, 3) composition of basic measures into components with greater dispersion. While time patterns of activity are not obvious, participant activity follow concentrations expected by scale-free networks. Comparison with ideal Erd\"os-R\'enyi network with the same number of edges and vertexes revealed as a sound criterion  for distinguishing sectors on the networks. Principal components in basic measures spaces revealed interesting and regular patterns of independence and dispersion. This includes a ranking of measures that most contribute to dispersion: 1) degree and strength measures, 2) symmetry related quantization, and 3) clusterization. Results suggested typologies for these networks and participants. Further work include considerations of text production, psychoanalysis inspired typologies, participatory democracy exploitation of observed properties, and better visualization support for network evolution.
\end{abstract}

@@

Text production and typologies of online participants benefit from results here presented~\cite{rcText,rcTipo}. 
Although all networks considered originated from email lists,
 coherence with literature suggests that results hold for a more general class of interaction networks,
 such as observed in online platforms (LinkedIn, Facebook, Twitter).

@@

 Such results are in accordance with phenomena observed in this work and linguistic characterization is being described in a deriving article~\cite{rcText}.

@@

\begin{figure}[hb]
    \centering
    \includegraphics[width=0.5\textwidth]{figs2/criaRede_}
    \caption{Formation of interaction network from email messages. Each vertex represents a participant. If participant B replyes participant A, that is regarded as evidence that B received information from A. Multiple messages add ``weight'' to directed edge. Further details are in Section~\ref{intNet}.}
    \label{formationNetwork}
\end{figure}

@@

\begin{figure}[hb]
    \centering
    \includegraphics[width=0.5\textwidth]{figs2/fser}
    \caption{Degree distribution on scale-free and Erd\"os-R\'enyi ideal networks. The later has more
        intermediary, as the former has more peripheral and hubs. Sections are
        given by the two intersections of the distributions $k_1$ and $k_2$. Characteristic degrees
    are in compact intervals of degree: $[0,k_1]$, $(k_1,k_2]$, $(k_2,k_{max}]$ for the three sections considered (periphery, intermediary and hubs).}
    \label{fig:setores}
\end{figure}

@@

        \subsubsection{Prevalence of topological measures}
This article
restricts this analisys to a small selection of the most basic
measures of each vertex:

@@

        \subsubsection*{Visualization of network evolution}
In refining hypotheses, visualization of the network was crucial. Animations, image galleries and an online gadgets were made~\cite{animacoes,galGMANE,appGMANE}. Mapping of various topological measures to glyphs and layouts are being further explored as a parallel research. Furthermore, stable aspects of measures prominance along time are captured through mean and standard deviation (see Section~\ref{}). Constant sector sizes along time are observed in a timeline fashion in Appendix~\ref{}.

@@

\section{Results and discussion}
    \subsection{Constancy and discrepancy of activity along time}\label{constDisc}
        \subsubsection{Seconds and minutes}
The incidence of messages at each second in a minute and at each minute in an hour is compatible with uniform distribution tests.
If compared to simulations using an uniform distribution\footnote{Numpy version 1.6.1, ``random.randint'' function.}, messages were slightly more evenly distributed in all lists: for both seconds and minutes  $\frac{max(incidence)}{min(incidence)} \in (1.26,1.275]$. Simulations reach these values, but have in average more discrepant higher and lower peaks $\xi=\frac{max(incidence')}{min(incidence')} \Rightarrow \mu_\xi=1.2918 \text{ and } \sigma_\xi=0.04619$ (see subsection~\ref{consistency}).

@@

    \subsection{Prevalence of centrality over asymmetries and asymetries over clusterization}\label{measures}

@@

\begin{figure} 
   \centering
        \includegraphics[width=\columnwidth]{figs2/ev0pr3PCA}
    \caption{PCA of in and out degree and strength, betweenness centrality and clustering coefficient, as specified in Section~\ref{measures}. Table~\ref{compPCA} has the composition of principle components from the original measures. First principle component is a pondered sum of centrality measures: degrees, strengthes and betweenness centrality. Second component is mostly clustering coefficient in this figure, but asymetries holds second component if also considered. Similarity to plot in Figure~\ref{clust} was verified with all window sizes considered ($ws\;\in\;[100,10000]$), which exposes a common relation is held by degree, strength and betweenness measures to clustering coefficient.}
    \label{PCA}
\end{figure}

\begin{figure} 
   \centering
        \includegraphics[width=\columnwidth]{figs2/ev0pr1PCA}
    \caption{Degree and strength, clustering coefficient, betweenness centrality and symmetry related measures are used for this scatter plot of principal components. Compositions of first three components are in table~\ref{compPCA2} and measure details in subsection~\ref{measures}. Most importantly, clustering coefficient is only relevant for third component, being second component representative of symmetry measurements of vertex interactions.}
    \label{PCA2}
\end{figure}

\begin{figure} 
   \centering
        \includegraphics[width=\columnwidth]{figs2/ev0pr11CC}
    \caption{Clustering versus degree of vertex in interaction network observed, $ws = 1000$ email messages, LAU list. General layout is in accordance with literature. With ver connected vertexes with low clusterization and high clusterization being gradually more incident as number of connections is lowered.}
    \label{clust}
\end{figure}

@@

\begin{acknowledgments}
Renato Fabbri is grateful to CNPq (process: 140860/2013-4,
project 870336/1997-5), United Nations Development Program (PNUD/ONU, contract: 2013/000566; project BRA/12/018)  and 
the Postgraduate Committee of the IFSC/USP. This author is also grateful for
the American Jewish Committee for maintaining an online copy of the Adorno book
used on the epigraph~\cite{adorno}. Authors thanks Gmane creators and keepers, specifically: Gmane is run by Lars Magne Ingebrigtsen, and the administrators are Tom Koelman, Jason R. Mastaler, Steinar Bang, Jon Ericson, Wolfgang Schnerring, Sebastian D.B. Krause, Nicolas Bareil, Raymond Scholz, and Adam Sjøgren. Authors thank refered email lists communities and welcome feedback as core contribution to this, and similar, research.
\end{acknowledgments}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

\section{Tables}\label{sectables}
\subsection{PCA tables}

\begin{table}
  \centering
  \begin{tabular}{|l|c|c| c|c| c|c|}\hline
 & \multicolumn{2}{c|}{PC1}          & \multicolumn{2}{c|}{PC2} & \multicolumn{2}{c|}{PC3}  \\\hline
       & $\mu$            & $\sigma$ & $\mu$         & $\sigma$ & $\mu$ & $\sigma$  \\\hline
$d$       & {\bf 48.02}   & 1.39     & 2.82          & 1.74     & 48.09  & 0.32 \\
$cc$      & 4.12          & 2.94     & {\bf 90.45}   & 3.98     & 3.98  & 0.77 \\ 
$bt$      & {\bf 47.87}   & 1.55     & 6.74          & 4.08     & 47.93 & 0.46 \\ \hline
$\lambda$ & 64.67         & 0.52     & 33.26         & 0.23     & 2.08  & 0.40 \\ \hline
  \end{tabular}
  \caption{Principal components composition in the simplest case: with degree, clustering coefficient and betweenness centrality. LAU list, $ws=1000$ messages in 20 disjoint positioning was used for statistics. First component is a pondered sum of degree and betweenness centrality measures. Second component is mostly clustering coefficient. First and second components sum more than 95\% of total dispersion.}
  \label{compPCA0}
\end{table}

@@

@@

\section{Data and scripts further description}\label{scripts}
Messages are downloaded from GMANE database by RSS in the mbox email text format. 
They are requested one by one to avoid reaching maximum size of the requests accepted by
GMANE API.

\subsection{Python scripts}\label{ap:os}
All results were obtained with scripts writen in the Python programming language. These are kept in a public repository for backup and sharing with research community~\cite{scriptsFim}. Core scripts, for deriving structures and results exhibited in this article, are in the LEIAME file.

\section{Figures of vertex classification fractions as the network evolves}\label{figures}

@@

\begin{figure*}[hb]
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/10000}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp10000}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/10000_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{benchmarks}.}
    \label{fig:cpp10000_}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/5000}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp5000}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/5000_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{benchmarks}.}
    \label{fig:cpp5000_}
\end{figure*}

%%%%
\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/1000}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp1000}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/1000_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{benchmarks}.}
    \label{fig:cpp1000_}
\end{figure*}

%%%
\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/500}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp500}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/500_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{benchmarks}.}
    \label{fig:cpp500_}
\end{figure*}

%%%%%%%%%
\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/250}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp250}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/250_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{benchmarks}.}
    \label{fig:cpp250_}
\end{figure*}

%%%%%%%
\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/100}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp100}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/100_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{benchmarks}.}
    \label{fig:cpp100_}
\end{figure*}

%%%%%%%%%%
\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/50}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp50}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/CPP/50_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{benchmarks}.}
    \label{fig:cpp50_}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs2/LAD/10000}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:lad10000}
\end{figure*}

>>>>

%\author{Luciano da Fontoura Costa}
%  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
%  \email{ldfcosta@gmail.com}
%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%
%\author{Osvaldo N. Oliveira Jr.}
%  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
%  \email{chu@ifsc.usp.br}
% \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\

@@

\begin{abstract}
    This article reports a characterization of interaction networks and its stability. Such a task involves a selection of aspects to investigate, which lead to: 1) activity distribution in time and among participants, 2) a sound classification of vertex: peripheral, intermediary and hub sectors, 3) combination of basic measures into components with greater dispersion. While time patterns of activity are not obvious, participant activity follows a scale-free distribution. Comparison with ideal Erd\"os-R\'enyi network with the same number of edges and vertexes was a sound criterion  for distinguishing sectors on the networks. Principal components in basic measures spaces revealed interesting and regular patterns of independence and dispersion. This includes a ranking of measures that most contribute to dispersion: 1) degree and strength measures, 2) symmetry related quantization, and 3) clusterization. Results suggest typologies for these networks and participants. Further work include considerations of text production, psychoanalysis inspired typologies, participatory democracy exploitation of observed properties, and better visualization support for network evolution.
\end{abstract}

@@

Text mining and typologies of online participants benefit from results here presented~\cite{rcText,rcTipo}. 
Although all networks considered originated from email lists,
 coherence with literature suggests that results hold for a more general class of interaction networks,
 such as observed in online platforms (LinkedIn, Facebook, Twitter).

@@

 Such results are in accordance with phenomena observed in this work and linguistic characterization is being described in a deriving article~\cite{rcText}. See Appendix~\ref{sec:fure} for further considerations of related work.

@@

\begin{figure}[hb]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/criaRede_}
    \caption{Formation of interaction network from email messages. Each vertex represents a participant. If participant B replyes participant A, that is regarded as evidence that B received information from A. Multiple messages add ``weight'' to directed edge. Further details are in Section~\ref{intNet}.}
    \label{formationNetwork}
\end{figure}

@@

\begin{figure}[hb]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/fser}
    \caption{Degree distribution on scale-free and Erd\"os-R\'enyi ideal networks. The later has more
        intermediary, as the former has more peripheral and hubs. Sections are
        given by the two intersections of the distributions $k_1$ and $k_2$. Characteristic degrees
    are in compact intervals of degree: $[0,k_1]$, $(k_1,k_2]$, $(k_2,k_{max}]$ for the three sections considered (periphery, intermediary and hubs).}
    \label{fig:setores}
\end{figure}

@@

        \subsubsection{Topological measures}\label{measures}
This article
restricts this analisys to a small selection of the most basic
measures of each vertex:

@@

        \subsubsection*{Visualization of network evolution}
In refining hypotheses, visualization of the network was crucial. Animations, image galleries and an online gadgets were made~\cite{animacoes,galGMANE,appGMANE}. Mapping of various topological measures to glyphs and layouts are being further explored as a parallel research. Furthermore, stable aspects of measures prominance along time are captured through mean and standard deviation (see Section~\ref{measures}). Constant sector sizes along time are observed in a timeline fashion in Appendix~\ref{figures}.

@@

\section{Results and discussion}
    \subsection{Constancy and discrepancy of activity along time}\label{constDisc}
        \subsubsection{Seconds and minutes}
The incidence of messages at each second in a minute and at each minute in an hour is compatible with uniform distribution tests.
If compared to simulations using an uniform distribution\footnote{Numpy version 1.6.1, ``random.randint'' function.}, messages were slightly more evenly distributed in all lists: for both seconds and minutes  $\frac{max(incidence)}{min(incidence)} \in (1.26,1.275]$. Simulations reach these values, but have in average more discrepant higher and lower peaks $\xi=\frac{max(incidence')}{min(incidence')} \Rightarrow \mu_\xi=1.2918 \text{ and } \sigma_\xi=0.04619$.

@@

    \subsection{Prevalence of centrality over asymmetries and asymetries over clusterization}\label{prevalence}

@@

\begin{figure} 
   \centering
        \includegraphics[width=\columnwidth]{figs/ev0pr3PCA}
    \caption{PCA of in and out degree and strength, betweenness centrality and clustering coefficient, as specified in Section~\ref{measures}. Table~\ref{compPCA} has the composition of principle components from the original measures. First principle component is a pondered sum of centrality measures: degrees, strengthes and betweenness centrality. Second component is mostly clustering coefficient in this figure, but asymetries holds second component if also considered. Similarity to plot in Figure~\ref{clust} was verified with all window sizes considered ($ws\;\in\;[100,10000]$), which exposes a common relation is held by degree, strength and betweenness measures to clustering coefficient.}
    \label{PCA}
\end{figure}

\begin{figure} 
   \centering
        \includegraphics[width=\columnwidth]{figs/ev0pr1PCA}
    \caption{Degree and strength, clustering coefficient, betweenness centrality and symmetry related measures are used for this scatter plot of principal components. Compositions of first three components are in table~\ref{compPCA2} and measure details in subsection~\ref{measures}. Most importantly, clustering coefficient is only relevant for third component, being second component representative of symmetry measurements of vertex interactions.}
    \label{PCA2}
\end{figure}

\begin{figure} 
   \centering
        \includegraphics[width=\columnwidth]{figs/ev0pr11CC}
    \caption{Clustering versus degree of vertex in interaction network observed, $ws = 1000$ email messages, LAU list. General layout is in accordance with literature. With ver connected vertexes with low clusterization and high clusterization being gradually more incident as number of connections is lowered.}
    \label{clust}
\end{figure}

@@

\begin{acknowledgments}
Renato Fabbri is grateful to CNPq (process: 140860/2013-4,
project 870336/1997-5), United Nations Development Program (PNUD/ONU, contract: 2013/000566; project BRA/12/018)  and 
the Postgraduate Committee of the IFSC/USP. This author is also grateful for
the American Jewish Committee for maintaining an online copy of the Adorno book
used on the epigraph~\cite{adorno}. Authors thanks Gmane creators and maintainers, specifically: Gmane is run by Lars Magne Ingebrigtsen, and the administrators are Tom Koelman, Jason R. Mastaler, Steinar Bang, Jon Ericson, Wolfgang Schnerring, Sebastian D.B. Krause, Nicolas Bareil, Raymond Scholz, and Adam Sjøgren. Authors thank refered email lists communities and welcome feedback as core contribution to this, and similar, research.
\end{acknowledgments}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Data and scripts further description}\label{scripts}
Messages are downloaded from GMANE database by RSS in the mbox email text format. 
They are requested one by one to avoid reaching maximum size of the requests accepted by
GMANE API.

\subsection{Python scripts}\label{ap:os}
All results were obtained with scripts writen in the Python programming language. These are kept in a public repository for backup and sharing with research community~\cite{scriptsFim}. Core scripts, for deriving structures and results exhibited in this article, are in the LEIAME file.
\section{Further consideration of related work}\label{sec:fure}
Unreciprocated edges often exceed 50\%, which matches empirical evidence reported in~\cite{newmanEvolving}. Although no correlation of topological characteristics and geographical position was found in a pertinent study~\cite{barabasiGeo}, geographical incidences should be present in further refinement of the analysis.

\section{Tables}\label{sectables}
\clearpage
\subsection{PCA tables}

\begin{table}[H]
  \centering
  \begin{tabular}{|l|c|c| c|c| c|c|}\hline
 & \multicolumn{2}{c|}{PC1}          & \multicolumn{2}{c|}{PC2} & \multicolumn{2}{c|}{PC3}  \\\hline
       & $\mu$            & $\sigma$ & $\mu$         & $\sigma$ & $\mu$ & $\sigma$  \\\hline
$d$       & {\bf 48.02}   & 1.39     & 2.82          & 1.74     & 48.09  & 0.32 \\
$cc$      & 4.12          & 2.94     & {\bf 90.45}   & 3.98     & 3.98  & 0.77 \\ 
$bt$      & {\bf 47.87}   & 1.55     & 6.74          & 4.08     & 47.93 & 0.46 \\ \hline
$\lambda$ & 64.67         & 0.52     & 33.26         & 0.23     & 2.08  & 0.40 \\ \hline
  \end{tabular}
  \caption{Principal components composition in the simplest case: with degree, clustering coefficient and betweenness centrality. LAU list, $ws=1000$ messages in 20 disjoint positioning was used for statistics. First component is a pondered sum of degree and betweenness centrality measures. Second component is mostly clustering coefficient. First and second components sum more than 95\% of total dispersion.}
  \label{compPCA0}
\end{table}

@@

\clearpage

@@

\clearpage
\section{Figures of vertex classification fractions as the network evolves}\label{figures}

@@

\begin{figure*}[hb]
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/10000}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp10000}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/10000_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{sectioning}.}
    \label{fig:cpp10000_}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/5000}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp5000}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/5000_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{sectioning}.}
    \label{fig:cpp5000_}
\end{figure*}

%%%%
\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/1000}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp1000}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/1000_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{sectioning}.}
    \label{fig:cpp1000_}
\end{figure*}

%%%
\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/500}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp500}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/500_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{sectioning}.}
    \label{fig:cpp500_}
\end{figure*}

%%%%%%%%%
\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/250}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp250}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/250_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{sectioning}.}
    \label{fig:cpp250_}
\end{figure*}

%%%%%%%
\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/100}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp100}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/100_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{sectioning}.}
    \label{fig:cpp100_}
\end{figure*}

%%%%%%%%%%
\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/50}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:cpp50}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/CPP/50_2}
    \caption{Distribution of vertex with respect to each measure of activity: compound criteria. Red, green and blue are for hubs, intermediary and border (peripheral) vertex fractions. The first two plots picture classifications that are not functions. Thus, in the first plot, the fraction of vertexes with unique classification in plotted in black. On the second plot, black is used to depict how much a vertex was classified in more than section: $\frac{\text{number of classifications} - \text{number of nodes}}{\text{number of nodes}}$. Compound criteria described in subsection~\ref{sectioning}.}
    \label{fig:cpp50_}
\end{figure*}

\begin{figure*}[hbtp] 
   \centering
        \includegraphics[width=\textwidth]{figs/LAD/10000}
    \caption{Distribution of vertex with respect to each measure of activity: in and out degrees and strengths. CPP Std library official mailing list. In the first six plots, red is fraction of hubs, green is the fraction of intermediary and blue is for peripheral fraction. On the last plot, red is the center (maximum distance to another vertex in equal to radius), blue is periphery (maximum distance equals to diameter) of the giant component. On the same graph, green counts the disconnected vertexes.}
    \label{fig:lad10000}
\end{figure*}
